---
title: "Explore Human BioMolecular Atlas Program Data Portal"
author: 
  - name: "Christine Hou"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: chris2018hou@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
package: "HuBMAPR"
vignette: |
  %\VignetteIndexEntry{Accessing Human Cell Atlas Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

'HuBMAP' data portal (<https://portal.hubmapconsortium.org/>) provides an open,
global bio-molecular atlas of the human body at the cellular level. 
`HuBMAPR` package provides an alternative interface to access the data
exploration and file retrieval.

# Installation

`HuBMAPR` is a R package available in *Bioconductor* version 3.19 and later.
You can install `HuBMAPR` by using the following commands in R session 
from *Bioconductor*:

```{r 'install', eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("HuBMAPR")

## Check Bioconductor installation
BiocManager::valid()
```

Additionally, you can install development version from
[GitHub](https://christinehou11.github.io/HuBMAPR):

```{r 'install dev', eval = FALSE}
BiocManager::install("christinehou11/HuBMAPR")
```

# Basic User Guide

## Load Necessary Packages

Install additional required packages before running package codes.

```{r 'pkg install', eval=FALSE}

```

Load additional packages. `dplyr` package is widely used in this vignettes to
conduct data wrangling and specific information extraction.

```{r 'library', message=FALSE}
library(dplyr)
library(tidyr)
library(HuBMAPR)
```

## Data Discovery

`HuBMAP` data portal page displays five categories of entity data, which are
**Dataset, Sample, Donor, Publication, and Collection**, chronologically (last
modified date time). You can choose to explore arbitrary size of data starting
from random location. By default, you can only page through less than 10000
data at one time using `from` and `size` numeric parameters. To read next or
more pages, change `from` to other values. Using corresponding functions to
explore entity data.

```{r 'main functions'}
# Retrieve the first 200 datasets
datasets_sub <- datasets(size = 200, from = 0)
datasets_sub

# Retrieve 200 samples from the 30th sample arranged chronologically
samples_sub <- samples(size = 200, from = 10)
samples_sub

# Retrieve full
datasets <- datasets()
samples <- samples()

donors <- donors()
donors

collections <- collections()
collections

publications <- publications()
publications
```

The default tibble produced by corresponding entity function only reflects
selected information. To see the names of selected information, use following
commands for each entity category. Specify `as` parameter to display
information in the format of `"character"` or `"tibble"`.

```{r 'cols'}
# as = "tibble" (default)
datasets_col_tbl <- datasets_default_columns(as = "tibble")
datasets_col_tbl

# as = "character"
datasets_col_char <- datasets_default_columns(as = "character")
datasets_col_char
```

`samples_default_columns(as)`, `donors_default_columns(as)`, 
`collections_default_columns(as)`, and `publications_default_columns(as)` 
work same as above.

A brief overview of selected information for five entity categories is:

```{r 'summary cols'}
tbl <- bind_cols(
    dataset = datasets_default_columns(as = "character"),
    sample = c(samples_default_columns(as = "character"), rep(NA, 6)),
    donor = c(donors_default_columns(as = "character"), rep(NA, 5)),
    collection = c(collections_default_columns(as = "character"),
                    rep(NA, 9)),
    publication = c(publications_default_columns(as = "character"), 
                    rep(NA, 6))
)

tbl
```

Use `organ()` to read through the available organs included in `HuBMAP`. It can
be helpful to filter retrieved data based on organ information.

```{r 'organs'}
organs <- organ()
organs
```

Data wrangling and filter are welcome to retrieve data based on interested
information.

```{r 'datasets filter'}
# Example from datasets()
datasets |>
    filter(organ == 'Small Intestine') |>
    count()
```

Any dataset, sample, donor, collection, and publication has special 
**HuBMAP ID** and **UUID**, and **UUID** is the main ID to be used in most of 
functions for specific detail retrievals. 

Since `datasets(size, from)`, `samples(size, from)`, `donors(size, from)`,
`collections(size, from)`, and `publications(size, from)` returned selected
information, you can use `*_detail(uuid)` to retrieve all available information
for any entry of any entity category given **UUID**. Use `select()` and
`unnest_*()` functions to expand list-columns.

```{r '*_detail()'}
dataset_uuid <- datasets |>
    filter(dataset_type == "Auto-fluorescence",
            organ == "Kidney (Right)") |>
    head(1) |>
    pull(uuid)

# Full Information
dataset_detail(dataset_uuid)

# Specific Information
dataset_detail(uuid = dataset_uuid) |>
    select(contributors) |>
    unnest_longer(contributors) |>
    unnest_wider(everything())
```

`sample_detail(uuid)`, `donor_detail(uuid)`, `collection_detail(uuid)`, and
`publication_detail(uuid)` work same as above.

## Derived Data

Some datasets from **Data** entity has derived (support) dataset(s). Use 
`dataset_derived(uuid)` to retrieve. The error message will be printed out if
the given dataset UUID has no derived/support dataset; otherwise, selected 
details will be retrieved as a tibble.

```{r 'dataset derived'}
# no derived/support dataset
dataset_uuid_1 <- "3acdb3ed962b2087fbe325514b098101"

dataset_detail(uuid = dataset_uuid_1)

# has derived/support dataset
dataset_uuid_2 <- "baf976734dd652208d13134bc5c4594b"

dataset_detail(uuid = dataset_uuid_2)
```

**Sample** and **Donor** have derived samples and datasets. In `HuBAMPR`
package, `sample_derived(uuid)` and `donor_derived(uuid)` functions are
available to use to see the derived datasets and samples from one sample given
sample UUID or one donor given donor UUID. Specify `entity_type` parameter to
retrieve derived `Dataset` or `Sample`.

```{r 'derived using sample_derived'}
sample_uuid <- samples |>
    filter(last_modified_timestamp >= "2023-01-01" &
            last_modified_timestamp <= "2023-10-01",
            organ == "Kidney (Left)") |>
    head(1) |>
    pull(uuid)

sample_uuid

# Derived Datasets
sample_derived(uuid = sample_uuid, entity_type = "Dataset")

# Derived Samples
sample_derived(uuid = sample_uuid, entity_type = "Sample")
```

`donor_derived(uuid)` works same as above.

In the retrieved tibbles using `samples(size, from)` and `donors(size, from)`,
there is one column named **donor.hubmap_id**. This can help to join the tibble
from `donors(size, from)` with tibbles from `datasets(size, from)` and
`samples(size, from)`.

```{r 'derived using left_join'}
donor_sub <- donors |>
    filter(Sex == "Female",
            Age <= 76 & Age >= 55,
            Race == "White",
            BMI <= 25,
            last_modified_timestamp >= "2020-01-08" &
            last_modified_timestamp <= "2020-06-30") |>
    head(1)

# Datasets
donor_sub_dataset <- donor_sub |>
    left_join(datasets |>
                select(-c(group_name, last_modified_timestamp)) |>
                rename("dataset_uuid" = "uuid",
                        "dataset_hubmap_id" = "hubmap_id"), 
                by = c("hubmap_id" = "donor.hubmap_id"))

donor_sub_dataset

# Samples
donor_sub_sample <- donor_sub |>
    left_join(samples |>
                select(-c(group_name, last_modified_timestamp)) |>
                rename("sample_uuid" = "uuid",
                        "sample_hubmap_id" = "hubmap_id"),
                by = c("hubmap_id" = "donor.hubmap_id"))

donor_sub_sample
```

The retrieved tibbles are same as 
`donor_derived(uuid, entity_type = "Dataset")` and 
`donor_derived(uuid, entity_type = "Sample")`.

```{r}
# Derived Datasets
donor_sub_uuid <- donor_sub |>
    pull(uuid)

donor_sub_uuid_dataset <- donor_derived(uuid = donor_sub_uuid,
                                        entity_type = "Dataset")

all(donor_sub_dataset$dataset_uuid %in% donor_sub_uuid_dataset$uuid)

# Derived Samples
donor_sub_uuid_sample <- donor_derived(uuid = donor_sub_uuid,
                                        entity_type = "Sample")

all(donor_sub_sample$sample_uuid %in% donor_sub_uuid_sample$uuid)

length(donor_sub_sample$sample_uuid) == 
        length(donor_sub_uuid_sample$uuid)
```


**Note:** For **Donor** entity, `left_join()` and `donor_derived(uuid)` produce
same results for both derived datasets and derived samples. For **Sample**
entity, `left_join()` and `sample_derived(uuid)` produce same results only for
derived datasets. *Some* samples are derived from another sample but they are
all derived from one donor. Therefore, we may see more results than that from
`sample_derived(uuid, entity_type = "Sample")`. Example shows below.

```{r 'example derived sample left_join'}
sample_donor_hubmap_id <- samples |>
    filter(uuid == sample_uuid) |>
    pull(donor.hubmap_id)

sample_donor_hubmap_id

# Derived Samples -> Different
sample_hubmap_id_sample_1 <- samples |>
    filter(donor.hubmap_id == sample_donor_hubmap_id)

sample_hubmap_id_sample_1

sample_hubmap_id_sample_2 <- sample_derived(uuid = sample_uuid, 
                                            entity_type = "Sample")

sample_hubmap_id_sample_2

length(sample_hubmap_id_sample_1$uuid) == 
        length(sample_hubmap_id_sample_2$uuid)


# Derived Datasets -> Same
sample_hubmap_id_dataset_1 <- datasets |>
    filter(donor.hubmap_id == sample_donor_hubmap_id)

sample_hubmap_id_dataset_2 <- sample_derived(uuid = sample_uuid,
                                            entity_type = "Dataset")

all(sample_hubmap_id_dataset_1$uuid %in%
        sample_hubmap_id_dataset_2$uuid)

length(sample_hubmap_id_dataset_1$uuid) == 
        length(sample_hubmap_id_dataset_2$uuid)
```

## Provenance Data

## Related Data

Each **Collection** has related datasets, and use `collection_data(uuid)` to
retrieve.

```{r 'collection datasets'}
collection_uuid <- collections |>
    filter(last_modified_timestamp >= "2023-01-01") |>
    head(1) |>
    pull(uuid)

collection_data(collection_uuid)
```


Each publication has related datasets, samples, and donors, and use
`publication_data(uuid)` to see, while specifying `entity_type` parameter to
retrieve derived `Dataset` or `Sample`.

```{r 'publication data'}
publication_uuid <- publications |>
    filter(publication_venue == "Nature") |>
    head(1) |>
    pull(uuid)

publication_data(publication_uuid, entity_type = "Dataset")
```

## Additional Information

To read the textual description of one **Collection** or **Publication**, use
`collection_information(uuid)` or `publication_information(uuid)` respectively.

```{r 'information'}
collection_information(uuid = collection_uuid)

publication_information(uuid = publication_uuid)
```

Some additional contact/author/contributor information can be retrieved using
`collection_contact(uuid)` and `collection_contributors(uuid)` for
**Collection** entity, or `publication_authors(uuid)` for **Publication**
entity.

```{r 'author'}
# Collection
collection_contacts(uuid = collection_uuid)

collection_contributors(uuid = collection_uuid)

# Publication
publication_authors(uuid = publication_uuid)
```

# File Download

For each dataset, there are corresponding data files. Most of the datasets'
files are saved in Globus HuBMAP Public Collection, and the URL is available.
Some of the datasets' files are not available via Globus, but can be accessed
via dbGAP (database of Genotypes and Phenotypes) and/or SRA (Sequence Read
Archive). But some of the datasets' files are not available in any authorized
platform. 

Use `files_globus_url(uuid)` to access data files. If the Globus URL is 
available, the function will direct you to chrome, and you can choose to 
download or read any document. If not, error messages will be printed out with
addition instructions, either providing dbGAP/SRA URLs or pointing out the
unavailability.

```{r 'file retrieval', eval=FALSE}
uuid_globus <- "d1dcab2df80590d8cd8770948abaf976"

files_globus_url(uuid_globus)

uuid_dbGAP_SRA <- "d926c41ac08f3c2ba5e61eec83e90b0c"

files_globus_url(uuid_dbGAP_SRA)

uuid_not_avail <- "0eb5e457b4855ce28531bc97147196b6"

files_globus_url(uuid_not_avail)
```

# `R` session information {.unnumbered}

```{r 'sessionInfo', echo=FALSE}
## Session info
options(width = 120)
sessionInfo()
```
