---
title: "Explore Human BioMolecular Atlas Program Data Portal"
author: 
  - name: "Christine Hou"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: chris2018hou@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
package: "HuBMAPR"
vignette: |
  %\VignetteIndexEntry{Accessing Human Cell Atlas Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Overview

'HuBMAP' data portal (https://portal.hubmapconsortium.org/) provides an open, global bio-molecular atlas of the human body at the cellular level. `HuBMAPR` package provides an alternative interface to access the data exploration and file retrieval.

# Installation

`HuBMAPR` is a R package available in *Bioconductor* version 3.19 and later. You can install `HuBMAPR` by using the following commands in R session from *Bioconductor*:

```{r 'install', eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("HuBMAPR")

## Check Bioconductor installation
BiocManager::valid()
```

Additionally, you can install development version from [GitHub](https://christinehou11.github.io/HuBMAPR):

```{r 'install dev', eval = FALSE}
BiocManager::install("christinehou11/HuBMAPR")
```

# Basic User Guide

## Load Necessary Packages

Install additional required packages before running package codes, and you can use

```{r, eval=FALSE}

```

Load additional packages. `dplyr` package is widely used in this vignettes to conduct data wrangling and specific information extraction.

```{r, message=FALSE, comment=NA}
library(tidyr)
library(HuBMAPR)
```

## Data Discovery

`HuBMAP` data portal page displays five categories of entity data (datasets, samples, donors, publications, and collections) chronologically (last modified date time). You can choose to explore arbitrary size of data starting from random location. By default, you can only page through less than 10000 data at one time using `from` and `size` numeric parameters. To read next or more pages, change `from` to other values. Using corresponding functions to explore entity data.

### Datasets

Retrieve the first 200 datasets using `datasets()`.

```{r, comment=NA}
datasets(size = 200, from = 0)
```




### Samples

### Donors

### Publications

### Collections


## File Retieval


```{r date, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

# `R` session information {.unnumbered}

```{r sessionInfo, echo=FALSE}
## Session info
options(width = 120)
sessionInfo()
```
