---
title: "Explore Human BioMolecular Atlas Program Data Portal"
author:
  - name: "Christine Hou"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: chris2018hou@gmail.com
output:
  BiocStyle::html_document
package: "HuBMAPR"
vignette: |
  %\VignetteIndexEntry{Accessing Human Cell Atlas Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

'HuBMAP' data portal (<https://portal.hubmapconsortium.org/>) provides an
open, global bio-molecular atlas of the human body at the cellular level.
`HuBMAPR` package provides an alternative interface to explore the data via R.

# Installation

`HuBMAPR` is a R package. Install development version from
[GitHub](https://christinehou11.github.io/HuBMAPR):

```{r 'install dev', eval = FALSE}
remotes::install("christinehou11/HuBMAPR")
```

# Basic User Guide

## Load Necessary Packages

Install additional required packages before running package codes.

```{r 'pkg install', eval=FALSE}
pkgs <- c("dplyr", "tidyr")
required_pkgs <- pkgs[!pkgs %in% rownames(installed.packages())]
BiocManager::install(required_pkgs)
```

Load additional packages. `dplyr` package is widely used in this vignettes to
conduct data wrangling and specific information extraction.

```{r 'library', message=FALSE}
library(dplyr)
library(tidyr)
library(HuBMAPR)
```

## Data Discovery

`HuBMAP` data portal page displays five categories of entity data, which are
**Dataset, Sample, Donor, Publication, and Collection**, chronologically (last
modified date time). Using corresponding functions to explore entity data.

```{r 'main functions'}
datasets <- datasets()
datasets

samples <- samples()
samples

donors <- donors()
donors

collections <- collections()
collections

publications <- publications()
publications
```

The default tibble produced by corresponding entity function only reflects
selected information. To see the names of selected information, use following
commands for each entity category. Specify `as` parameter to display
information in the format of `"character"` or `"tibble"`.

```{r 'cols'}
# as = "tibble" (default)
datasets_col_tbl <- datasets_default_columns(as = "tibble")
datasets_col_tbl

# as = "character"
datasets_col_char <- datasets_default_columns(as = "character")
datasets_col_char
```

`samples_default_columns()`, `donors_default_columns()`,
`collections_default_columns()`, and `publications_default_columns()` work same
as above.

A brief overview of selected information for five entity categories is:

```{r 'summary cols'}
tbl <- bind_cols(
    dataset = datasets_default_columns(as = "character"),
    sample = c(samples_default_columns(as = "character"), rep(NA, 7)),
    donor = c(donors_default_columns(as = "character"), rep(NA, 6)),
    collection = c(collections_default_columns(as = "character"),
                    rep(NA, 10)),
    publication = c(publications_default_columns(as = "character"),
                    rep(NA, 7))
)

tbl
```

Use `organ()` to read through the available organs included in `HuBMAP`. It can 
be helpful to filter retrieved data based on organ information.

```{r 'organs'}
organs <- organ()
organs
```

Data wrangling and filter are welcome to retrieve data based on interested
information.

```{r 'datasets filter'}
# Example from datasets()
datasets |>
    filter(organ == 'Small Intestine') |>
    count()
```

Any dataset, sample, donor, collection, and publication has special 
**HuBMAP ID** and **UUID**, and **UUID** is the main ID to be used in most of
functions for specific detail retrievals.

The column of **donor.hubmap_id** is included in the retrieved tibbles from
`samples()` and `datasets()`, which can help to join the tibble.

```{r 'derived using left_join'}
donor_sub <- donors |>
    filter(Sex == "Female",
            Age <= 76 & Age >= 55,
            Race == "White",
            `Body Mass Index` <= 25,
            last_modified_timestamp >= "2020-01-08" &
            last_modified_timestamp <= "2020-06-30") |>
    head(1)

# Datasets
donor_sub_dataset <- donor_sub |>
    left_join(datasets |>
                select(-c(group_name, last_modified_timestamp)) |>
                rename("dataset_uuid" = "uuid",
                        "dataset_hubmap_id" = "hubmap_id"),
                by = c("hubmap_id" = "donor.hubmap_id"))

donor_sub_dataset

# Samples
donor_sub_sample <- donor_sub |>
    left_join(samples |>
                select(-c(group_name, last_modified_timestamp)) |>
                rename("sample_uuid" = "uuid",
                        "sample_hubmap_id" = "hubmap_id"),
                by = c("hubmap_id" = "donor.hubmap_id"))

donor_sub_sample
```

You can use `*_detail(uuid)` to retrieve all available information for any entry
of any entity category given **UUID**. Use `select()` and `unnest_*()` functions
to expand list-columns. It will be convenient to view tables with multiple 
columns but one row using `glimpse()`.

```{r '*_detail()'}
dataset_uuid <- datasets |>
    filter(dataset_type == "Auto-fluorescence",
            organ == "Kidney (Right)") |>
    head(1) |>
    pull(uuid)

# Full Information
dataset_detail(dataset_uuid) |> glimpse()

# Specific Information
dataset_detail(uuid = dataset_uuid) |>
    select(contributors) |>
    unnest_longer(contributors) |>
    unnest_wider(everything())
```

`sample_detail()`, `donor_detail()`, `collection_detail()`, and
`publication_detail()` work same as above.

## Metadata

To retrieve the metadata for **Dataset**, **Sample**, and **Donor** metadata, 
use `dataset_metadata()`, `sample_metadata()`, and `donor_metadata()`.

```{r 'metadata'}
dataset_metadata("993bb1d6fa02e2755fd69613bb9d6e08")

sample_metadata("8ecdbdc3e2d04898e2563d666658b6a9")

donor_metadata("b2c75c96558c18c9e13ba31629f541b6")
```

## Derived Data

Some datasets from **Dataset** entity has derived (support) dataset(s). Use
`dataset_derived()` to retrieve. A tibble with selected details will be 
retrieved as if the given dataset has support dataset; otherwise, nothing 
returns.

```{r 'dataset derived'}
# no derived/support dataset
dataset_uuid_1 <- "3acdb3ed962b2087fbe325514b098101"

dataset_derived(uuid = dataset_uuid_1)

# has derived/support dataset
dataset_uuid_2 <- "baf976734dd652208d13134bc5c4594b"

dataset_derived(uuid = dataset_uuid_2) |> glimpse()
```

**Sample** and **Donor** have derived samples and datasets. In `HuBAMPR` 
package, `sample_derived()` and `donor_derived()` functions are available to use
to see the derived datasets and samples from one sample given sample UUID or one
donor given donor UUID. Specify `entity_type` parameter to retrieve derived
`Dataset` or `Sample`.

```{r 'derived using sample_derived'}
sample_uuid <- samples |>
    filter(last_modified_timestamp >= "2023-01-01" &
            last_modified_timestamp <= "2023-10-01",
            organ == "Kidney (Left)") |>
    head(1) |>
    pull(uuid)

sample_uuid

# Derived Datasets
sample_derived(uuid = sample_uuid, entity_type = "Dataset")

# Derived Samples
sample_derived(uuid = sample_uuid, entity_type = "Sample")
```

`donor_derived()` works same as above.

## Provenance Data

For individual entries from **Dataset** and **Sample** entities,
`uuid_provenance()` helps to retrieve the provenance of the entry as a list of
characters (UUID, HuBMAP ID, and entity type) from the most recent ancestor to 
the furthest ancestor. There is no ancestor for Donor UUID, and an empty list 
will be returned.

```{r 'provenance'}
# dataset provenance
dataset_uuid <- "3e4c568d9ce8df9d73b8cddcf8d0fec3"
uuid_provenance(dataset_uuid)

# sample provenance
sample_uuid <- "35e16f13caab262f446836f63cf4ad42"
uuid_provenance(sample_uuid)

# donor provenance
donor_uuid <- "0abacde2443881351ff6e9930a706c83"
uuid_provenance(donor_uuid)
```

## Related Data

Each **Collection** has related datasets, and use `collection_data()` to 
retrieve.

```{r 'collection datasets'}
collection_uuid <- collections |>
    filter(last_modified_timestamp >= "2023-01-01") |>
    head(1) |>
    pull(uuid)

collection_data(collection_uuid)
```

Each publication has related datasets, samples, and donors, and use
`publication_data()` to see, while specifying `entity_type` parameter to 
retrieve derived `Dataset` or `Sample`.

```{r 'publication data'}
publication_uuid <- publications |>
    filter(publication_venue == "Nature") |>
    head(1) |>
    pull(uuid)

publication_data(publication_uuid, entity_type = "Dataset")

publication_data(publication_uuid, entity_type = "Sample")
```

## Additional Information

To read the textual description of one **Collection** or **Publication**, use
`collection_information()` or `publication_information()` respectively.

```{r 'information'}
collection_information(uuid = collection_uuid)

publication_information(uuid = publication_uuid)
```

Some additional contact/author/contributor information can be retrieved using
`dataset_contributor()` for **Dataset** entity, `collection_contact()` and
`collection_contributors()` for **Collection** entity, or 
`publication_authors()` for **Publication** entity.

```{r 'author'}
# Dataset
dataset_contributors(uuid = dataset_uuid)

# Collection
collection_contacts(uuid = collection_uuid)

collection_contributors(uuid = collection_uuid)

# Publication
publication_authors(uuid = publication_uuid)
```

# File Download

For each dataset, there are corresponding data files. Most of the datasets' 
files are available on HuBMAP Globus with corresponding URL. Some of the 
datasets' files are not available via Globus, but can be accessed via dbGAP
(database of Genotypes and Phenotypes) and/or SRA (Sequence Read Archive). But
some of the datasets' files are not available in any authorized platform.

Use `bulk_data_transfer()` to know whether data files are open-accessed or
restricted. The function will direct you to chrome if the files are accessible;
otherwise, the error messages will be printed out with addition instructions,
either providing dbGAP/SRA URLs or pointing out the unavailability.

```{r 'bulk data transfer', eval=FALSE}
uuid_globus <- "d1dcab2df80590d8cd8770948abaf976"

bulk_data_transfer(uuid_globus)

uuid_dbGAP_SRA <- "d926c41ac08f3c2ba5e61eec83e90b0c"

bulk_data_transfer(uuid_dbGAP_SRA)

uuid_not_avail <- "0eb5e457b4855ce28531bc97147196b6"

bulk_data_transfer(uuid_not_avail)
```

You can choose to download data files on Globus webpage by clicking download
button after choosing the desired document. You can also preview and download 
data files using [rglobus](https://mtmorgan.github.io/rglobus/) package.
Follow the instructions 
[here](https://mtmorgan.github.io/rglobus/articles/a_get_started.html).

# `R` session information {.unnumbered}

```{r 'sessionInfo', echo=FALSE}
## Session info
options(width = 120)
sessionInfo()
```
